Polynomial Time and Dependent Types

### 2023-10-24 paper/paper.tex
[ paper ] tidy up

Type theory is often claimed to be ideal for Computer Science,
combining programming and proof in one unifying system, so a happy
programmer can verify while they program, and program while they
verify. From a broader Computer Science view, however, type theory
lacks the ability to talk about the very thing that makes Computer
Science interesting -- the fact that computation is everywhere bounded
by the resources in time and space that we can afford to give it.

Typical Type Theory only speaks of the public face that programs
present to the world -- if you input things like this, you get things
that look like that -- but cannot bring itself to mention the true
cost of programs' execution. One can encode costs by embedding another
programming language in type theory, for example \cite{GueneauCP18},
or one can synthesise costs by treating resource counting as a
computational effect, for example \cite{Danielsson08,NiuSGH22}, but
neither of these capture the intrinsic costs of the programs we write
in type theory. These techniques deliver only \emph{conspicuous
  consumption}, not speaking of the real resources consumed.

In this paper, we propose a method for extending dependent type theory
with a means for constraining the intrinsic computational complexity
of programs written in the theory. We concentrate on linear type
systems that soundly and completely capture polynomial time
computation, the commonly used standard for feasible resource usage,
and extend these systems to dependent types. The additional
expressivity of dependent types allows us use these characterisations
of polytime to further functionally characterise the classes of
non-deterministic and probabilistic polynomial time.

@@

Our long-term goal is to combine the extensional reasoning of Type
Theory with intensional reasoning about the resources intrinsically
consumed by programs. This paper is a first step along this path,
which we hope will lead both to practical systems for reasoning about
programs' resource usage, and to theoretical use as a form of
\emph{synthetic computational complexity theory}. We discuss these
possibilities further in \autoref{sec:conclusion}.

% The goal of this paper is to unify implicit computational complexity
% with full dependent types. ICC enables us to ensure that
% \emph{programs} are within a certain complexity class, while dependent
% types allow us to state and prove that these programs solve the
% problems that we intend them to, and to construct reductions between
% problems. We can marry complexity-controlled implementations with
% problems, paving the way for a \emph{synthetic computational
%   complexity theory}.

% \item Section 5: Extensions
%   \begin{enumerate}
%   \item Explicit Complexity via Diamond Counting
%   \item Poly-time univalence
%   \item Towards P- and NP-completeness proofs. Issue here is the
%     extensionality. Towards a synthetic (poly-time based)
%     computational complexity theory.
%   \end{enumerate}
% \item Section 6: Related Work
%   \begin{enumerate}
%   \item ICC stuff
%     \begin{enumerate}
%     \item LFPL
%     \item CONS-free
%     \item Soft Affine Logic; which has the same underlying idea as
%       CONS-free (is this a novel observation?)
%     \item Other systems based on information flow.
%     \end{enumerate}
%   \item Explicit complexity: lambda-amor, and things by Deepak Garg
%     and Marco Gaboardi.
%   \item Recent CMU stuff on calf
%   \end{enumerate}

@@

This paper makes the following contributions to the theory and use of
linear dependent type theory and implicit computational complexity:
\begin{enumerate}
\item We formulate two systems that combine linear type theory for
  polytime computation with full dependent types, using Quantitative
  Type Theory. The systems are presented in \autoref{sec:qtt}. The
  linear typing discipline required for enforcing polytime is provided
  by QTT, but we also need to carefully add constructs for
  non-iterable datatypes (\autoref{sec:noniter-qtt}) and the two kinds
  of natural number iterator that we consider
  (\autoref{sec:cons-free-qtt} and \autoref{sec:lfpl-qtt}). Porting
  the natural number iterators from the simply typed to the
  dependently typed setting requires careful annotation of the rules
  to ensure that the correct information is available for type
  checking, while also not allowing too much information to be made
  available at runtime that would violate the polytime soundness
  property. A further contribution of this paper is the addition of
  reflection types to QTT, \autoref{sec:qtt-reflection}, which allow
  statements about polytime realisability to be reflected into
  types. % We explain in \autoref{Sec:realisability-future} how this
  % will have applications in QTT beyond the polytime systems presented
  % here.
\item We demonstrate the utility of the combination of polytime and
  dependent types in \autoref{sec:programming-polytime}. Just as in
  the simply typed world, we have an expressive language for writing
  polytime programs. With the additional power of dependent types, we
  can also prove properties of these programs. A simple example is
  proving that a sorting program actually sorts. Using QTT reflection,
  we can go further and represent the class of polytime problems, with
  polytime reductions between them, as dependent pairs
  (\autoref{sec:polytime-problems}). Our final examples use dependent
  types to give monadic presentations of the complexity classes of
  Non-deterministic Polynomial (NP) time and Probabilistic Polynomial
  (PP) time. Since these classes rely on specific correctness
  criteria, it is not possible to capture them in a simply typed
  system for polytime.
\item We prove the polytime soundness of our systems via a
  realisability argument in \autoref{sec:soundness}. Our construction
  is an extension of the amortised complexity realisability
  constructions of \citet{dallago11realisability}. We extend their
  work to our dependently typed setting, and also give a realisability
  interpretation of datatypes directly, instead of via second-order
  impredicative encodings. The technical content of this section has
  been formalised in the Agda proof assistant
  \cite{norell2008dependently}, and is included in the supplementary
  material.
\end{enumerate}

@@

% \subsection{Explicit Resource Tracking}

% \begin{enumerate}
% \item We still require diamonds to build zero and successor, but the
%   iterator does not give them back.
% \item In the simply typed case, we can annotate types with the static
%   amount of potential they have per constructor.
% \end{enumerate}

@@

% \begin{displaymath}
%   \begin{array}{l}
%     \mathrm{runWithOracle} \stackrel0: \mathrm{ND}\,A \to \ListTy(\BoolTy) \to \mathrm{Maybe}\,A \\
%     \mathrm{runWithOracle}\,(\mathsf{return}\,a)\,\mathit{bs} = \mathsf{just}\,a \\
%     \mathrm{runWithOracle}\,(\mathsf{choice}\,x\,y)\,(\cTrue :: \mathit{bs}) = \mathrm{runWithOracle}\,x\,\mathit{bs} \\
%     \mathrm{runWithOracle}\,(\mathsf{choice}\,x\,y)\,(\cFalse :: \mathit{bs}) = \mathrm{runWithOracle}\,y\,\mathit{bs} \\
%     \mathrm{runWithOracle}\,(\mathsf{choice}\,x\,y)\,\cNil = \mathsf{nothing}
%   \end{array}
% \end{displaymath}

@@

The differencing operator of the natural number resource monoid can
only supply as much fuel as is contained in the potential.
% This makes
% it suitable for amortised analysis, where we directly store potential
% in data structures.
For the two polynomial time systems, we need more sophisticated
structures, both originally presented by Dal Lago and Hofmann. The
fundamental idea with both is to represent potentials as pairs
$(m,p)$, where $m$ is a natural number and $p$ is a polynomial. The
$m$ tracks the ``size'' of data as it pertains to the number of times
an operation will be repeated by iterating over it --- for example, an
iterable natural number will have size equal to itself, but a
non-iterable natural number may be assigned zero size. The polynomial
$p$ tracks the complexity of a program as a function of the size of
the input. This leads to a differencing operator that evaluates the
polynomial with the size of the data:

@@

% FIXME: difference between this definition and Hofmann-Dal Lago's: we
% use polynomials with natural number coefficients instead of a function
% that is bounded by a polynomial.

@@

% \subsubsection{Realising QTT}

% Summing up the previous sections, we have the following key theorem
% paving the way to polytime soundness:
% \begin{theorem}\label{thm:basic-qtt-realisability-model}
%   For every resource monoid with a distinguished sub-monoid
%   $(M, M_0)$, we obtain a realisability model of $\mathbb{N}$-QTT with
%   $\Pi$ and $\Sigma$ types, a universe, extensional equality types,
%   realisability reflection, and non-iterable datatypes.
% \end{theorem}

>>>>

Type Theory is often claimed to be ideal for Computer Science,
combining programming and proof in one unifying system, so a happy
programmer can verify while they program, and program while they
verify. From a broader Computer Science view, however, Type Theory
lacks the ability to talk about the very thing that makes Computer
Science interesting -- the fact that computation is everywhere bounded
by the resources in time and space that we can afford to give it.

Typical Type Theory only speaks of the public face that programs
present to the world -- if you input things like this, you get things
that look like that -- but cannot bring itself to mention the true
cost of programs' execution. One can encode costs by embedding another
programming language in Type Theory, for example \cite{GueneauCP18},
or one can synthesise costs by treating resource counting as a
computational effect, for example \cite{Danielsson08,NiuSGH22}, but
neither of these capture the intrinsic costs of the programs we write
in Type Theory. These techniques deliver only \emph{conspicuous
  consumption}, not speaking of the real resources consumed.

In this paper, we propose a method for extending dependent Type Theory
with a means for constraining the intrinsic computational complexity
of programs written in the theory. We concentrate on linear type
systems that soundly and completely capture polynomial time
computation, the commonly used standard for feasible resource usage,
and extend these systems to dependent types. The additional
expressivity of dependent types allows us use these characterisations
of polytime to further functionally characterise the classes of
non-deterministic and probabilistic polynomial time.

@@

Our long-term goal is to combine the extensional reasoning of Type
Theory with intensional reasoning about the resources intrinsically
consumed by programs. This paper is a first step along this path,
which we hope will lead both to practical systems for reasoning about
programs' resource usage as well as their extensional behaviour, and
to theoretical use as a form of \emph{synthetic computational
  complexity theory}. We discuss these possibilities further in
\autoref{sec:conclusion}.

@@

This paper makes the following contributions to the theory and use of
linear dependent type theory and implicit computational complexity:
\begin{enumerate}
\item We formulate two systems that combine linear type theory for
  polytime computation with full dependent types, using Quantitative
  Type Theory. The systems are presented in \autoref{sec:qtt}. The
  linear typing discipline required for enforcing polytime is provided
  by QTT, but we also need to carefully add constructs for
  non-iterable datatypes (\autoref{sec:noniter-qtt}) and the two kinds
  of natural number iterator that we consider
  (\autoref{sec:cons-free-qtt} and \autoref{sec:lfpl-qtt}). Porting
  the natural number iterators from the simply typed to the
  dependently typed setting requires careful annotation of the rules
  to ensure that the correct information is available for type
  checking, while also not allowing too much information to be made
  available at runtime that would violate the polytime soundness
  property. A further contribution of this paper is the addition of
  reflection types to QTT, \autoref{sec:qtt-reflection}, which allow
  statements about polytime realisability to be reflected into
  types.
\item We demonstrate the utility of the combination of polytime and
  dependent types in \autoref{sec:programming-polytime}. Just as in
  the simply typed world, we have an expressive language for writing
  polytime programs. With the additional power of dependent types, we
  can also prove properties of these programs. A simple example is
  proving that a sorting program actually sorts. Using QTT reflection,
  we can go further and represent the class of polytime problems, with
  polytime reductions between them, as dependent pairs
  (\autoref{sec:polytime-problems}). Our final examples use dependent
  types to give monadic presentations of the complexity classes of
  Non-deterministic Polynomial (NP) time and Probabilistic Polynomial
  (PP) time. Since these classes rely on specific correctness
  criteria, it is not possible to capture them in a simply typed
  system for polytime.
\item We prove the polytime soundness of our systems via a
  realisability argument in \autoref{sec:soundness}. Our construction
  is an extension of the amortised complexity realisability
  constructions of \citet{dallago11realisability}. We extend their
  work to our dependently typed setting, and also give a realisability
  interpretation of datatypes directly, instead of via second-order
  impredicative encodings. The technical content of this section has
  been formalised in the Agda proof assistant
  \cite{norell2008dependently}, and is included in the supplementary
  material.
\end{enumerate}

@@

@@

@@

The differencing operator of the natural number resource monoid can
only supply as much fuel as is contained in the potential.  For the
two polynomial time systems, we need more sophisticated structures,
both originally presented by Dal Lago and Hofmann. The fundamental
idea with both is to represent potentials as pairs $(m,p)$, where $m$
is a natural number and $p$ is a polynomial. The $m$ tracks the
``size'' of data as it pertains to the number of times an operation
will be repeated by iterating over it --- for example, an iterable
natural number will have size equal to itself, but a non-iterable
natural number may be assigned zero size. The polynomial $p$ tracks
the complexity of a program as a function of the size of the
input. This leads to a differencing operator that evaluates the
polynomial with the size of the data:

@@

@@
